---
title: "mini_project_1"
format: html
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(mdsr)      
library(maps)
library(viridis)
library(leaflet)
library(sf)
```

##Dataset: CigarettesSW: This dataset is about cigarette consumption for the 48 continental US States from 1985–1995.

##Source: Stock, J.H. and Watson, M.W. (2007). Introduction to Econometrics, 2nd ed. Boston: Addison Wesley.

##Website: https://vincentarelbundock.github.io/Rdatasets/datasets.html

```{r}
# Load US map data
library(maps)
us_states <- map_data("state")
head(us_states)
```

```{r}
#| message: false
#| warning: false

# Load CigarettesSW dataset
CigarettesSW <- read.csv("https://raw.githubusercontent.com/Ela-Kanade/CigarettesMaps/main/CigarettesSW.csv")

# Filter for 1985 data
cigarettes_data <- CigarettesSW %>%
  filter(year == 1985)

# Create a short version of the US map data (only one entry per region)
us_states_short <- us_states %>%
  group_by(region) %>%
  slice(1)
```

#Set up the CigarettesSW dataset!

```{r}
# Map state abbreviations to full names
state_abbr_to_fullname <- data.frame(
  abbr = state.abb,  
  name = state.name)

# Convert the 'region' column in us_states_short to title case for consistency
us_states_short$region <- str_to_title(us_states_short$region)  

# Join the state abbreviation data to cigarettes_data and rename column for consistency
cigarettes_data <- cigarettes_data %>%
  left_join(state_abbr_to_fullname, by = c("state" = "abbr")) %>%
  rename(states_names = name)

# Merge the cleaned US states data with cigarettes data
combined_data <- us_states_short %>%
  left_join(cigarettes_data, by = c("region" = "states_names"))

us_map <- map_data("state")
```

#Static plot of Cigarette Packs Sold by State!

```{r}
# Create a static map of cigarette packs sold by state
library(ggplot2)

# Prepare the map data and add cigarette packs sold per state
state_packs <- combined_data %>%
  select(region, packs) %>%
  mutate(region = tolower(region))  # Match lowercase region names in map_data

# Merge the map data with cigarette packs data
map_with_packs <- left_join(us_map, state_packs, by = "region")

# Create the plot
ggplot(data = map_with_packs, aes(x = long, y = lat, group = group, fill = packs)) +
  geom_polygon(color = "black", linewidth = 0.2) +
  scale_fill_viridis(option = "D", na.value = "gray90") +
  theme_void() +
  labs(title = "Cigarette Packs Sold by State", fill = "Packs Sold")
```


#More set up for interactive plots

```{r}
#| warning: false
# Convert long and lat to numeric and ensure packs are numeric as well
combined_data$long <- as.numeric(combined_data$long)
combined_data$lat <- as.numeric(combined_data$lat)
combined_data$packs <- as.numeric(combined_data$packs)

# Convert to an sf object for spatial manipulation
us_states_sf <- st_as_sf(combined_data, coords = c("long", "lat"), crs = 4326, agr = "constant")
```

# First interactive plot - numerical

```{r}
#| warning: false
# Load spatial data for US states
states_sf <- read_sf("https://rstudio.github.io/leaflet/json/us-states.geojson") %>%
  select(name, geometry)

# Merge spatial data with cigarette data
cigarettes_sf <- states_sf %>%
  left_join(cigarettes_data, by = c("name" = "states_names")) %>%
  filter(!(name %in% c("alaska", "hawaii"))) %>%
  mutate(across(where(is.numeric), ~round(.x, 1)))

# Create color bins for the numerical data (packs sold per capita)
bins <- c(0, 30, 60, 90, 120, 150, 200, 250, Inf)
pal <- colorBin("YlOrRd", domain = cigarettes_sf$packs, bins = bins)
```


```{r}
#| warning: false
# Add labels to each state and create the interactive leaflet map
library(htmltools)
library(glue)

# Add labels for each state with the number of packs sold
cigarettes_sf <- cigarettes_sf %>%
  mutate(labels = str_c(name, ": ", packs, " packs of cigarettes sold per capita in 1985"))

# Convert labels to HTML format for leaflet
labels <- lapply(cigarettes_sf$labels, HTML)

# Create the interactive map with leaflet
leaflet(cigarettes_sf) %>%
  setView(-96, 37.8, 4) %>%
  addTiles() %>%
  addPolygons(
    fillColor = ~pal(packs),
    weight = 2,
    opacity = 1,
    color = "black",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) %>%
  addLegend(pal = pal, values = ~packs, opacity = 0.7, title = NULL, position = "bottomright")
```

# Second interactive plot - categorical

```{r}
# Categorize cigarette sales by packs sold
cigarettes_sf <- states_sf %>%
  left_join(cigarettes_data, by = c("name" = "states_names")) %>%
  filter(!(name %in% c("alaska", "hawaii"))) %>%
  mutate(across(where(is.numeric), ~round(.x, 1))) %>%
  mutate(cigarette_category = case_when(
    packs < 90 ~ "Low",
    packs >= 91 & packs < 110 ~ "Medium",
    packs >= 111 & packs < 130 ~ "High",
    packs >= 130 ~ "Very High"
  ))

# Add labels for each state with the number of packs sold
cigarettes_sf <- cigarettes_sf %>%
  mutate(labels = str_c(name, ": ", packs, " packs of cigarettes sold per capita in 1985"))

# Convert labels to HTML format for leaflet
labels <- lapply(cigarettes_sf$labels, HTML)

# Create color palette for categorical data
factpal <- colorFactor(c("lightblue", "cyan2", "mediumturquoise", "darkblue"), 
                       levels = c("Low", "Medium", "High", "Very High"))

# Create the interactive map with leaflet
leaflet(cigarettes_sf) %>%
  setView(-96, 37.8, 4) %>%
  addTiles() %>%
  addPolygons(
    weight = 2,
    opacity = 1,
    color = "black",
    fillColor = ~factpal(cigarette_category),  # Color by the categorical variable
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 5,
      color = "maroon",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) %>%
  addLegend(
    pal = factpal, 
    values = ~cigarette_category,  # Use the categorical values for the legend
    opacity = 0.7, 
    title = NULL, 
    position = "bottomright")
```
